---
title: "Project-G1"
author: "Group 1"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(ggplot2) # for plotting a bit easier
# #install.packages("VGAM")
# library(VGAM)
# #install.packages("distr6")
# library(distr6)
library(ggthemes)
theme_set(theme_pander())
```

# Group project, initial report  

### Queueing system in observation  

Group 1 data collected from Wellington Train Station the terminal station for Wellington city and the southern-most terminal of passenger trains in the North Isand of New Zealand.  The queueing system in observation is train station passenger ticketing booths.  The ticketing booths provide sales and services for the public and passengers on the trains.  The queueing system is located adjacent to a supermarket, cafe, bar, and is en-route between the passenger platforms and the Wellington City main bus terminal. Customers can enter the queueing system:  

  - after disembarking a passenger train  
  - as pedestrian traffic from several street level and underground walkways  
  
The Wellington train station ticketing booths form a M/M/C system, where C (the number of service channels) ranges between 3 and 4.  Data was collected over four sessions across four separate dates.  Each observation records the date and time of arrival in the system, time of entering service, and time of departing service.  From these data, the inter-arrival times and service times are calculated (as seconds), and from these the mean inter-arrival and service times are calculated.  


###  Reading in the data  

```{r read in data}
train.data <- read.csv("ProjectData-G1.csv")

#add factors to sessions and servers
train.data$Session <- factor(train.data$Session)
train.data$Servers <- factor(train.data$Servers)
train.data$Lane <- factor(train.data$Lane)

# scale the inter arrivals and service times
train.data$arr_scaled <- scale(train.data$inter.arr.sec)
train.data$serv_scaled <- scale(train.data$Serv_time_sec)

#m majority of data 
train.data <- train.data[train.data$inter.arr.sec<=200,]
# objects to bin interarrival data
n_bins <- 100


str(train.data)
```

The structure of the dataset is presented here and comprises `r dim(train.data)` rows and columns.  


### Inter-arrival Time  

Inter-arrival rate is calculated as the sum of all inter-arrival times divided by the number of inter-arrival times.

```{r calculate lambda}
# estimate parameters for density plot, assuming exponential distribution
lamb <- mean(train.data$inter.arr.sec)
lamb
```

The inter-arrival times are plotted here, showing a histogram of the data, density function of the actual data and the theoretical exponential distribution, assuming $\lambda =$ `r lamb`.

### Plotting the inter arrival times  

```{r plot interarrival times as histogram, echo=FALSE, warning=FALSE}
# length of vectors for plotting
n_train <- length(train.data$inter.arr.sec)

library(ggplot2)
set.seed(300212228)

# plot inter-arrival times actual density and theoretical density
arr_dist_200 <- ggplot(train.data, aes(x = inter.arr.sec, )) +
  geom_histogram(aes(y = ..density..), 
                 bins = n_bins) + 
  geom_line(aes(y = ..density.., colour = "Sample"), 
            stat = "density", 
            size =1) + 
  stat_function(aes(x = rexp(n_train), colour = "Exp(1/lambda)"),
                fun = dexp, 
                args = list(rate = 1/lamb),) +
  
  stat_function(aes(x = rgamma(n_train, shape = 2, rate = 1/lamb),
                    colour = "Gamma(2, 1/lambda)"),
                fun = dgamma, 
                args = list(shape = 2, rate = 1/lamb),
                ) +  
  stat_function(aes(x = rgamma(n_train, shape = 2, rate = 2/lamb),
                    colour = "Gamma(2, 2/lambda)"),
                fun = dgamma, 
                args = list(shape = 2, rate = 2/lamb),
                ) +    
  stat_function(aes(x = rgamma(n_train, shape = 2, rate = 3/lamb),
                    colour = "Gamma(2, 3/lambda)"),
                fun = dgamma, 
                args = list(shape = 2, rate = 3/lamb),
                ) +    
  stat_function(aes(x = rgamma(n_train, shape = 3, rate = 2/lamb),
                    colour = "Gamma(3, 2/lambda)"),
                fun = dgamma, 
                args = list(shape = 3, rate = 2/lamb),
                ) 

arr_dist_200 +
  labs(title = "Inter-arrival times histogram and density \n 0-200 seconds",
       x = "Inter-arrival time (s)",
       colour = "Distributions") +
  scale_x_continuous(limits = c(0, 200)) +
  scale_y_continuous(limits = c(0, 0.04))


```



## Goodness of fit   

The inter-arrival time visually looks like a good fit to the exponential distribution, so we will begin by running a goodness of fit test for the data and set our rejection level to $\alpha = 0.05$.  


```{r chisquare for interarrival time exponential distribution, echo=FALSE, warning=FALSE}
# establish intervals for bins
arr_ints <- round(seq(from = 0, to = 200, by = 200/n_bins))
arr_bins <- factor(findInterval(train.data$inter.arr.sec, arr_ints))

# create reference object to hold random variates from different distributions
# Exp(1/lamb), Gamma(2, 1/lamb), Gamma(2, 2/lamb)
set.seed(1234)
arr_exp_1lamb <- factor(round(rexp(n_train, rate = 1/lamb)))
arr_gam_2_1lamb <- factor(round(rgamma(n_train, shape = 2, rate = 1/lamb)))
arr_gam_2_2lamb <- factor(round(rgamma(n_train, shape = 2, rate = 2/lamb)))
arr_gam_2_3lamb <- factor(round(rgamma(n_train, shape = 2, rate = 3/lamb)))
arr_gam_3_2lamb <- factor(round(rgamma(n_train, shape = 3, rate = 2/lamb)))

# bin each object by interval
arr_exp_1_bin <- factor(findInterval(arr_exp_1lamb, arr_ints))
arr_gam_21_bin <- factor(findInterval(arr_gam_2_1lamb, arr_ints))
arr_gam_22_bin <- factor(findInterval(arr_gam_2_2lamb, arr_ints))
arr_gam_23_bin <- factor(findInterval(arr_gam_2_3lamb, arr_ints))
arr_gam_32_bin <- factor(findInterval(arr_gam_2_2lamb, arr_ints))

# store chisq.test results as object
arr_chi_e1 <- chisq.test(x = arr_bins, y = arr_exp_1_bin)
arr_chi_g21 <- chisq.test(x = arr_bins, y = arr_gam_21_bin)
arr_chi_g22 <- chisq.test(x = arr_bins, y = arr_gam_22_bin)
arr_chi_g23 <- chisq.test(x = arr_bins, y = arr_gam_23_bin)
arr_chi_g32 <- chisq.test(x = arr_bins, y = arr_gam_32_bin)


# display chisq.test results
arr_chi_e1
arr_chi_g21
arr_chi_g22
arr_chi_g23
arr_chi_g32

```

### Results of the chi-square test:  

##### Testing for $Exp(\frac {1}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Exp(\frac {1}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Exp(\frac {1}{\lambda})$ distribution  


Since $p=$ `r arr_chi_e1$p.value` `r if (arr_chi_e1$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (arr_chi_g22$p.value > 0.05) {"is"} else {"is not"}` consistent with data from an exponential distribution with the $Exp(\frac {1}{\lambda})$ distribution.  


##### Testing for $Gamma(2, \frac {1}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(2, \frac {1}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(2, \frac {1}{\lambda})$ distribution  


Since $p=$ `r arr_chi_g21$p.value`, `r if (arr_chi_g21$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (arr_chi_g22$p.value > 0.05) {"is"} else {"is not"}` consistent with the $Gamma(2, \frac {1}{\lambda})$ distribution.  


##### Testing for $Gamma(2, \frac {2}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(2, \frac {2}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(2, \frac {2}{\lambda})$ distribution  


Since $p=$ `r arr_chi_g22$p.value`, `r if (arr_chi_g22$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (arr_chi_g22$p.value > 0.05) {"is"} else {"is not"}`  consistent with the $Gamma(2, \frac {2}{\lambda})$ distribution.    


##### Testing for $Gamma(2, \frac {3}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(2, \frac {3}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(2, \frac {3}{\lambda})$ distribution  


Since $p=$ `r arr_chi_g23$p.value`, `r if (arr_chi_g23$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (arr_chi_g23$p.value > 0.05) {"is"} else {"is not"}`  consistent with the $Gamma(2, \frac {3}{\lambda})$ distribution.


##### Testing for $Gamma(3, \frac {2}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(3, \frac {2}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(3, \frac {2}{\lambda})$ distribution  


Since $p=$ `r arr_chi_g32$p.value`, `r if (arr_chi_g32$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (arr_chi_g23$p.value > 0.05) {"is"} else {"is not"}`  consistent with the $Gamma(3, \frac {2}{\lambda})$ distribution.


The highest p-value is `r max(arr_chi_g32$p.value, arr_chi_g23$p.value, arr_chi_g22$p.value, arr_chi_g21$p.value, arr_chi_e1$p.value)`
so we select the distribution:  

$Gamma(3, \frac {2}{\lambda})$
 

### Fitting a distribution  

Initial visual inspection of the plot shows that the data may fit a Gamma distribution with shape = 3, rate = $\frac {2}{\lambda}$.  This distribution is equivalent to an Erlang distribution with shape = 3, rate = $\frac {2}{\lambda}$.  


```{r plot interarrival times with best fit, echo = FALSE, warning=FALSE}
# plot inter-arrival times actual density and theoretical density
arr_dist_fin <- ggplot(train.data, aes(x = inter.arr.sec, )) +
  geom_histogram(aes(y = ..density..), 
                 bins = n_bins) + 
  geom_line(aes(y = ..density.., colour = "Sample"), 
            stat = "density", 
            size =1) + 
  scale_x_continuous(limits = c(0, 75)) +
  scale_y_continuous(limits = c(0, 0.04)) +
  stat_function(aes(x = rexp(n_train, rate = 1/lamb),
                    colour = "Exp(1/lambda)"),
                fun = dexp, 
                args = list(rate = 1/lamb),
                ) +    
  stat_function(aes(x = rgamma(n_train, shape = 2, rate = 2/lamb),
                    colour = "Gamma(2, 2/lambda)"),
                fun = dgamma, 
                args = list(shape = 2, rate = 2/lamb),
                ) +  
  stat_function(aes(x = rgamma(n_train, shape = 3, rate = 2/lamb),
                colour = "Gamma(3, 2/lambda)"),
                fun = dgamma, 
                args = list(shape = 3, rate = 2/lamb),
                ) + 
  labs(title = "Inter-arrival times histogram and density \n reduced 0 - 75 sec",
       x = "Inter-arrival time (s)",
       colour = "Distributions")

arr_dist_fin

```
 


## Service Time

The service time is calculated as the mean time a customer spends in service (i.e. does not include queueing time). `

```{r}
# calculate mean service time as mu
mu <- mean(train.data$Serv_time_sec)
mu
```

The service time is plotted here, showing a histogram of the service times, density plot of the actual data and theoretical density of the exponential distribution with $\mu=$ `r mu`.


```{r servtime plots, echo = FALSE}
# plot service times, actual density and theoretical density 
srv_dists <- ggplot(train.data, aes(x = Serv_time_sec, )) +
  geom_histogram(aes(y = ..density..), 
                 bins = n_bins, 
                 alpha = 0.4) + 
  geom_line(aes(y = ..density.., 
            colour = "Sample"), 
            stat = "density", 
            size = 1) + 

  stat_function(aes(x = rgamma(1:length(Serv_time_sec), shape = 4), 
                colour = "Gamma(4, 5/mu)"), 
                fun = dgamma, 
                args = list(shape= 4, rate = 5/mu)) +
  stat_function(aes(x = rexp(1:length(Serv_time_sec)),
                colour = "Exp(1/mu)"),
                fun = dexp,
                args = list(rate = 1/mu),) +
  stat_function(aes(x = rgamma(1:length(Serv_time_sec), shape = 3),
                colour = "Gamma(3, 4/mu)"),
                fun = dgamma,
                args = list(shape= 3, rate = 4/mu)) +

  stat_function(aes(x = rgamma(1:length(Serv_time_sec), shape = 4),
                colour = "Gamma(4, 4/mu)"),
                fun = dgamma,
                args = list(shape= 4, rate = 4/mu)) +

  stat_function(aes(x = rgamma(1:length(Serv_time_sec), shape = 5),
                colour = "Gamma(5, 5/mu)"),
                fun = dgamma,
                args = list(shape= 5, rate = 5/mu)) +
  stat_function(aes(x = rgamma(1:length(Serv_time_sec), shape = 6),
                colour = "Gamma(5, 6/mu)"),
                fun = dgamma,
                args = list(shape= 5, rate = 6/mu)) +
  stat_function(aes(x = rgamma(1:length(Serv_time_sec), shape = 6),
                colour = "Gamma(6, 5/mu)"),
                fun = dgamma,
                args = list(shape= 6, rate = 5/mu))

# display the plot
srv_dists  +
  labs(title = "Service times histogram and density \n 0-200 seconds",
       x = "Service time (s)",
       colour = "Distributions")


```



 
## Goodness of fit  
The service time visually looks like a possible fit to the exponential distribution, so we will begin by running a goodness of fit test for the data and set our rejection level to $\alpha = 0.05$.  


$H_0:$ The data is consistent with the Exponential distribution   
$H_1:$ The data is not consistent with the Exponential distribution  

```{r chisquare for service times, echo = FALSE, warning=FALSE}
# chi-square test for our data
# establish intervals for bins
srv_ints <- round(seq(from = 0, to = 200, by = 200/n_bins))
srv_bins <- factor(findInterval(train.data$Serv_time_sec, arr_ints))

# create reference object to hold random variates from different distributions
# Exp(1/lamb), Gamma(2, 1/lamb), Gamma(2, 2/lamb)
set.seed(1234)
srv_exp_1mu <- factor(round(rexp(n_train, rate = 1/mu)))
srv_gam_3_4mu <- factor(round(rgamma(n_train, shape = 3, rate = 4/mu)))
srv_gam_4_4mu <- factor(round(rgamma(n_train, shape = 4, rate = 4/mu)))
srv_gam_4_5mu <- factor(round(rgamma(n_train, shape = 4, rate = 5/mu)))
srv_gam_5_5mu <- factor(round(rgamma(n_train, shape = 5, rate = 5/mu)))
srv_gam_5_6mu <- factor(round(rgamma(n_train, shape = 5, rate = 6/mu)))
srv_gam_6_5mu <- factor(round(rgamma(n_train, shape = 6, rate = 5/mu)))


# bin each object by interval
srv_exp_1_bin <- factor(findInterval(srv_exp_1mu, srv_ints))
srv_gam_34_bin <- factor(findInterval(srv_gam_3_4mu, srv_ints))
srv_gam_44_bin <- factor(findInterval(srv_gam_4_4mu, srv_ints))
srv_gam_45_bin <- factor(findInterval(srv_gam_4_5mu, srv_ints))
srv_gam_55_bin <- factor(findInterval(srv_gam_5_5mu, srv_ints))
srv_gam_56_bin <- factor(findInterval(srv_gam_5_6mu, srv_ints))
srv_gam_65_bin <- factor(findInterval(srv_gam_6_5mu, srv_ints))

# store chisq.test results as object
srv_chi_e1 <- chisq.test(x = srv_bins, y = srv_exp_1_bin)
srv_chi_g34 <- chisq.test(x = srv_bins, y = srv_gam_34_bin)
srv_chi_g44 <- chisq.test(x = srv_bins, y = srv_gam_44_bin)
srv_chi_g45 <- chisq.test(x = srv_bins, y = srv_gam_45_bin)
srv_chi_g55 <- chisq.test(x = srv_bins, y = srv_gam_55_bin)
srv_chi_g56 <- chisq.test(x = srv_bins, y = srv_gam_56_bin)
srv_chi_g65 <- chisq.test(x = srv_bins, y = srv_gam_65_bin)

# display chisq.test results
srv_chi_e1
srv_chi_g34
srv_chi_g44
srv_chi_g45
srv_chi_g55
srv_chi_g56
srv_chi_g65

```


### Results of the chi-square test:  

##### Testing for $Exp(\frac {1}{\mu})$ distribution:  

$H_0:$ The data is consistent with the $Exp(\frac {1}{\mu})$ distribution   
$H_1:$ The data is not consistent with the $Exp(\frac {1}{\mu})$ distribution  


Since $p=$ `r srv_chi_e1$p.value` `r if (srv_chi_e1$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (srv_chi_e1$p.value > 0.05) {"is"} else {"is not"}` consistent with data from an exponential distribution with the $Exp(\frac {1}{\mu})$ distribution.  



##### Testing for $Gamma(3, \frac {4}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(3, \frac {4}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(3, \frac {4}{\lambda})$ distribution  


Since $p=$ `r srv_chi_g34$p.value`, `r if (srv_chi_g34$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (srv_chi_g34$p.value > 0.05) {"is"} else {"is not"}` consistent with the $Gamma(3, \frac {4}{\lambda})$ distribution.  



##### Testing for $Gamma(4, \frac {4}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(4, \frac {4}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(4, \frac {4}{\lambda})$ distribution  


Since $p=$ `r srv_chi_g44$p.value`, `r if (srv_chi_g44$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (srv_chi_g44$p.value > 0.05) {"is"} else {"is not"}`  consistent with the $Gamma(4, \frac {4}{\lambda})$ distribution.    



##### Testing for $Gamma(4, \frac {5}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(4, \frac {5}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(4, \frac {5}{\lambda})$ distribution  


Since $p=$ `r srv_chi_g45$p.value`, `r if (srv_chi_g45$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (srv_chi_g45$p.value > 0.05) {"is"} else {"is not"}`  consistent with the $Gamma(4, \frac {5}{\lambda})$ distribution.



##### Testing for $Gamma(5, \frac {5}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(5, \frac {5}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(5, \frac {5}{\lambda})$ distribution  


Since $p=$ `r srv_chi_g55$p.value`, `r if (srv_chi_g55$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (srv_chi_g55$p.value > 0.05) {"is"} else {"is not"}`  consistent with the $Gamma(5, \frac {5}{\lambda})$ distribution.



##### Testing for $Gamma(5, \frac {6}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(5, \frac {6}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(5, \frac {6}{\lambda})$ distribution  


Since $p=$ `r srv_chi_g56$p.value`, `r if (srv_chi_g56$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (srv_chi_g56$p.value > 0.05) {"is"} else {"is not"}`  consistent with the $Gamma(5, \frac {6}{\lambda})$ distribution.



##### Testing for $Gamma(6, \frac {5}{\lambda})$ distribution:  

$H_0:$ The data is consistent with the $Gamma(6, \frac {5}{\lambda})$ distribution   
$H_1:$ The data is not consistent with the $Gamma(6, \frac {5}{\lambda})$ distribution  


Since $p=$ `r srv_chi_g65$p.value`, `r if (srv_chi_g65$p.value > 0.05) {"is greater than 0.05 we fail to reject"} else {"is less than  0.05 we reject"}` the null hypothesis and conclude that the inter-arrival time `r if (srv_chi_g65$p.value > 0.05) {"is"} else {"is not"}`  consistent with the $Gamma(6, \frac {5}{\lambda})$ distribution.




The highest p-value is `r max(srv_chi_e1$p.value, srv_chi_g34$p.value, srv_chi_g44$p.value, srv_chi_g45$p.value, srv_chi_g55$p.value, srv_chi_g56$p.value, srv_chi_g65$p.value)`

so we select the distribution:  
  
$Gamma(5, \frac {6}{\lambda})$


```{r srvtimes focussed plot, echo = FALSE, warning=FALSE}
# display a focussed plot
srv_dist_75 <- srv_dists +  scale_x_continuous(limits = c(0, 75)) +
  scale_y_continuous(limits = c(0, 0.05))  +
  labs(title = "Service times histogram and density \n reduced scale 0 - 75 seconds",
       x = "Service time (s)",
       colour = "Distributions")
srv_dist_75

```



This is equivalent to an Erlang distribution with parameters $Erlang(5, \frac {6}{\mu})$.




```{r selected distribution for srvtimes, echo=FALSE}


# plot service times, actual density and theoretical density 
ggplot(train.data, aes(x = Serv_time_sec, )) +
  geom_histogram(aes(y = ..density..), 
                 bins = n_bins, 
                 alpha = 0.4) + 
  scale_x_continuous(limits = c(0, 75)) +
  scale_y_continuous(limits = c(0, 0.05))  +
  stat_function(aes(x = rexp(n_train),
                colour = "Exp(1/mu)"),
                fun = dexp,
                args = list(rate = 1/mu)) +
  stat_function(aes(x = rgamma(1:length(Serv_time_sec), shape = 5),
                colour = "Gamma(5, 6/mu)"),
                fun = dgamma,
                args = list(shape= 5, rate = 6/mu)) +
  geom_line(aes(y = ..density.., 
            colour = "Sample"), 
            stat = "density", 
            size = 1) + 
  labs(title = "Service times histogram and density \n reduced size 0-75 seconds",
       x = "Service time (s)",
       colour = "Densities")
  
  
```



### Conclusion:

Although two distributions share the same p-value for the _inter-arrival times best-fit_, we selected $Gamma(2, \frac {2}{\lambda})$ as the best-fit model.  This distribution is an $Erlang(2, \frac {2}{\lambda})$, a special case of the gamma distribution.

The best-fit mode for service times is $Exp\frac {1}{\mu})$.  


